---
title: "Project 3 Final"
author: "Lauren Urban"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary packages
```{r}
library(psych)
library(tidyverse)
library(lmtest)
library(gtsummary)
library(apaTables)
library(broom)
library(utils)
library(finalfit)
library(interactions)
library(randomizr)
library(caret)
```

# Reading in the data/summary statistics
```{r}
# Load CSV 
adpkd <- read.csv("/Users/laurenurban/Downloads/Project3_data.csv")

# View the first few rows 
head(adpkd)

# Summary statistics
str(adpkd)
summary(adpkd)

# Descriptive statistics
describe(adpkd)
```

# Descriptive statistics
```{r}
adpkd %>%
  summarize(
    # Geometry features
    mean_geom1 = mean(geom1, na.rm = TRUE),
    sd_geom1 = sd(geom1, na.rm = TRUE),
    mean_geom2 = mean(geom2, na.rm = TRUE),
    sd_geom2 = sd(geom2, na.rm = TRUE),

    # Gabor transform features
    across(starts_with("gabor"), 
           list(mean = ~ mean(.x, na.rm = TRUE), 
                sd = ~ sd(.x, na.rm = TRUE))),

    # Gray level co-occurrence matrix features
    across(starts_with("glcm"), 
           list(mean = ~ mean(.x, na.rm = TRUE), 
                sd = ~ sd(.x, na.rm = TRUE))),

    # Image texture features
    across(starts_with("txti"), 
           list(mean = ~ mean(.x, na.rm = TRUE), 
                sd = ~ sd(.x, na.rm = TRUE))),

    # Local binary pattern features
    across(starts_with("lbp"), 
           list(mean = ~ mean(.x, na.rm = TRUE), 
                sd = ~ sd(.x, na.rm = TRUE))),

    # Kidney volume measurements
    mean_tkvht_base = mean(tkvht_base, na.rm = TRUE),
    sd_tkvht_base = sd(tkvht_base, na.rm = TRUE),
    mean_tkvht_visit2 = mean(tkvht_visit2, na.rm = TRUE),
    sd_tkvht_visit2 = sd(tkvht_visit2, na.rm = TRUE),

    # Progression status (proportion of fast progression)
    progression_rate = mean(progression, na.rm = TRUE),

    # Percentage change in kidney volume
    mean_tkvht_change = mean(tkvht_change, na.rm = TRUE),
    sd_tkvht_change = sd(tkvht_change, na.rm = TRUE)
  )
```
# Separating data sets
```{r}
# Split into training and test sets
set.seed(123)  
trainIndex <- createDataPartition(adpkd$progression, p = 0.6, list = FALSE)
train_data <- adpkd[trainIndex, ]
test_data <- adpkd[-trainIndex, ]

# Check proportions of progression in each set to make sure they are ~ equal
train_proportion <- prop.table(table(train_data$progression))
test_proportion <- prop.table(table(test_data$progression))

cat("Proportion of progression in training set:\n")
print(train_proportion)
cat("Proportion of progression in test set:\n")
print(test_proportion)
```

# Squaring Baseline Kidney Volume
```{r}
# Square baseline kidney volume 
train_data <- train_data %>%
  mutate(tkvht_base_squared = tkvht_base^2)

test_data <- test_data %>%
  mutate(tkvht_base_squared = tkvht_base^2)
```

# Z-score normalize all variables
```{r}
zscore <- function(x, na.rm = T) {
    (x - mean(x, na.rm = T))/sd(x, na.rm = T)
}

train_data_z <- test_data |> 
  mutate(
    across(
      c(geom1:tkvht_base, tkvht_base_squared),
      zscore,
      .names = "z_{.col}"))
```

# Look at boxplots and t tests to identify features that differ by progression
```{r}

# Define feature names for the loop
feature_names <- c("geom1", "geom2", "gabor1", "gabor2", "gabor3", "tkvht_base", "tkvht_change")

# Loop through each feature and create a boxplot by progression group
for (feature in feature_names) {
  p <- ggplot(adpkd, aes(x = factor(progression), y = .data[[feature]])) +
    geom_boxplot() +
    labs(x = "Progression (0 = Slow, 1 = Fast)", y = feature) +
    ggtitle(paste("Boxplot of", feature, "by Progression Group"))
  
  print(p)  # This explicitly prints each plot
}

# Run t-tests for each feature and store results
t_test_results <- lapply(feature_names, function(feature) {
  t.test(adpkd[[feature]] ~ adpkd$progression)
})

# View results for each feature
t_test_results
```
# Cross validation
```{r}
# Randomly split data into 5 folds for cross-validation
control <- trainControl(
  method = "cv",
  number = 5
)
```

######## Task 1 #########
```{r}
# Model 1: Baseline height-corrected total kidney volume as predictor
model1 <- lm(tkvht_change ~ tkvht_base, data = train_data_z)

# Summarize the model to see the results
summary(model1)

# Make predictions on the test set
predictions1 <- predict(model1, newdata = test_data)

# View predictions
head(predictions1)

# Calculate RMSE (Root Mean Squared Error) for Model 1
rmse1 <- sqrt(mean((predictions1 - test_data$tkvht_change)^2))
cat("RMSE for Model 1: ", rmse1, "\n")
```
```{r}
# Model 2: Image features as predictors
model2 <- lm(tkvht_change ~ geom1 + geom2 + gabor1 + gabor4 + gabor5 + 
              glcm1 + txti2 + txti3 + lbp1 + lbp2 + lbp3 + lbp4 + 
              lbp5, data = train_data_z)

# Summarize the model to see the results
summary(model2)

# Based on p values, stepwise selecting covariates to include/remove from model
# Removing the following manually increases R squared
# gabor2, gclm2,txti5, txti1, txti4, gabor3
# txti3 remains despite its p value because model becomes less accurate when it is removed, all else are <0.05

# Make predictions on the test set
predictions2 <- predict(model2, newdata = test_data)

# View predictions
head(predictions2)

# Calculate RMSE (Root Mean Squared Error) for Model 1
rmse2 <- sqrt(mean((predictions2 - test_data$tkvht_change)^2))
cat("RMSE for Model 2: ", rmse2, "\n")
```
```{r}
# Model 3: Both baseline height-corrected total kidney volume and image features as predictors
model3 <- lm(tkvht_change ~ tkvht_base + geom1 + geom2 + gabor1 + gabor4 + gabor5 + 
              glcm1 + txti2 + txti3 + lbp1 + lbp2 + lbp3 + lbp4 + lbp5, 
              data = train_data_z)

# Summarize the model to see the results
summary(model3)

# Based on p values, stepwise selecting covariates to include/remove from model
# Removing the following manually increases R squared
# gabor2, txti5, gclm2, txti1, txti4, gabor3
# txti3 remains despite its p value because model becomes less accurate when it is removed, all else are <0.05

# Make predictions on the test set
predictions3 <- predict(model3, newdata = test_data)

# View predictions
head(predictions3)

# Calculate RMSE (Root Mean Squared Error) for Model 1
rmse3 <- sqrt(mean((predictions3 - test_data$tkvht_change)^2))
cat("RMSE for Model 3: ", rmse3, "\n")
```

# Model comparison for task 1:
```{r}
# Compare R-squared, Adjusted R-squared, AIC, and BIC for all models
# Model 1 (baseline height-corrected total kidney volume)
summary(model1)

# Model 2 (image features from MRI)
summary(model2)

# Model 3 (both baseline total kidney volume and image features)
summary(model3)

# Compare AIC and BIC
AIC(model1, model2, model3)
BIC(model1, model2, model3)

# Compare R-squared for all models
r_squared <- c(
  Model1 = summary(model1)$r.squared,
  Model2 = summary(model2)$r.squared,
  Model3 = summary(model3)$r.squared
)
r_squared
```

# Checking homoscedasticity
```{r}
# Breusch-Pagan test for homoscedasticity (constant variance of residuals)
bptest(model1)
bptest(model2)
bptest(model3)
```

# Check normality of residuals
```{r}
# Shapiro-Wilk test for normality of residuals
shapiro.test(resid(model1))
shapiro.test(resid(model2))
shapiro.test(resid(model3))
```

######### Task 2 #########
# Create and compare models using only height-corrected kidney volume at baseline
```{r}
prog_tkvht1 <- train(factor(progression) ~ z_tkvht_base,
                     data = train_data_z,
                     method = "regLogistic",
                     maximize = TRUE,
                     trControl = control
        )

prog_tkvht1 #best accuracy

prog_tkvht2 <- train(factor(progression) ~ z_tkvht_base_squared,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control
)

prog_tkvht2

prog_tkvht3 <- train(factor(progression) ~ z_tkvht_base + z_tkvht_base_squared,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control
)

prog_tkvht3
```

# Create and compare models using only image features
```{r}
prog_images1 <- train(factor(progression) ~ z_gabor2,
                     data = train_data_z,
                     method = "regLogistic",
                     maximize = TRUE,
                     trControl = control)

prog_images2 <- train(factor(progression) ~ z_txti2,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

prog_images3 <- train(factor(progression) ~ z_lbp3,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

prog_images4 <- train(factor(progression) ~ z_lbp5,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

models1 <- resamples(list(gabor = prog_images1, txti = prog_images2, lbp3 = prog_images3,
                          lbp5 = prog_images4))
summary(models1)

prog_images5 <- train(factor(progression) ~ z_gabor2 + z_txti2,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

prog_images6 <- train(factor(progression) ~ z_gabor2 + z_lbp3,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

prog_images7 <- train(factor(progression) ~ z_gabor2 + z_lbp5,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

models2 <- resamples(list(txti = prog_images5, lbp3 = prog_images6, lbp5 = prog_images7))
summary(models2)

prog_images8 <- train(factor(progression) ~ z_gabor2 + z_txti2 + lbp3,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

prog_images9 <- train(factor(progression) ~ z_gabor2 + z_txti2 + lbp5,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

models3 <- resamples(list(lbp3 = prog_images8, lbp5 = prog_images9))
summary(models3)

prog_images10 <- train(factor(progression) ~ z_gabor2 + z_txti2 + lbp3 + lbp5,
                      data = train_data_z,
                      method = "regLogistic",
                      maximize = TRUE,
                      trControl = control)

models4 <- resamples(list(lbp3 = prog_images8, lbp5 = prog_images10))
summary(models4)

models5 <- resamples(list(one = prog_images1, two = prog_images2, three = prog_images3,
                           four = prog_images4, five = prog_images5, six = prog_images6,
                           seven = prog_images7, eight = prog_images8, nine = prog_images9,
                           ten = prog_images10))
summary(models5)
```

